Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks

### 2022-10-31 rmd/manuscript.Rmd
edit

---
title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  pdf_document:
    latex_engine: xelatex
  #word_document:
  #  reference_docx: word_format.docx
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: ecosystems.csl
header-includes:
  \usepackage{textgreek}
---

@@

We employed a discrete-time model to describe tri-trophic dynamics in each habitat patch. A discrete model is suitable for organisms with predictable reproduction cycles, a common life history for many lotic species, including macroinvertebrates and fish. Our simulation model consists of three species; a basal species $B$ with positive logistic growth up to a carrying capacity, $K$; a primary consumer species or intraguild prey, $C$, which grows based on the consumption of $B$; and an intraguild predator $P$, which can consume both $B$ and $C$ (**Figure 1C**). In our model, the realized species abundance $A_{i,x}(t+1)$ for species i (either $B,C,~or~P$) at patch $x$ and time $t+1$ is expressed as a composite of within-patch trophic interactions, immigration, and emigration:

@@

where $s_{0,i}$ is the background survival probability (density independent), $\beta_{i,x}(t)$ the survival probability after predation, and $\gamma_{x}(t)$ the survival probability after stochastic disturbance. The function $g_i(\centerdot)$ defines the form of the density-dependent recruitment for species $i$.

For the basal species, we considered the predation impacts by consumer and predator species:

@@

$r_b$ is the reproductive rate for survived individuals.

@@

The interactive effect of disturbance and ecosystem structure has important implications for ecosystem management because anthropogenic climate change is predicted to increase the frequency and magnitude of disturbances [@brunnerFutureStreamflowRegime2020; @mittalImpactHumanIntervention2016]. River networks are especially vulnerable due to their propensity to go dry during droughts [@zipperPervasiveChangesStream2021] as well as susceptibility to more intense flood disturbance with increasing storm intensity [@schiermeierIncreasedFloodRisk2011]. Therefore, the preservation of habitat diversity conferred by ecosystem size and complexity is critical to biodiversity conservation under rapid environmental changes. In addition, maintaining habitat connectivity is particularly important for large, linearly-arranged networks. Although dispersal has received less attention in food chain research [but see @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012a; @wangMetapopulationCapacityDetermines2021], our simulation highlighted food webs in these networks are more vulnerable to habitat fragmentation because habitat patches are inherently isolated from one another. Dams and road crossings have altered the connectivity between suitable habitat patches or have completely blocked dispersal [@ishiyamaPredictingEcologicalImpacts2018; @moritaEffectsHabitatFragmentation2002; @perkinFragmentationAltersStream2012a]. Light pollution has also negatively affected the ability of adult aquatic insects to disperse between habitat patches [@perkinEffectsArtificialLighting2014]. These human activities may disproportionately impact food webs in large watersheds with simple network structures.

@@

Our findings add an important distinction to work to assess and prioritize protection or conservation status within human impacted regions. Classical conservation schemes emphasize ecosystem size, aiming to protect the largest areas possible [@diamondIslandDilemmaLessons1975]. While this is certainly a useful attribute generally, designating networks with a high degree of variability, or many small patches, can be more beneficial than a large, homogeneous area [@fahrigWhySeveralSmall2020]. Accounting for ecosystem complexity is particularly suited to ecosystem architectures in which branches have variable environmental conditions [@koningNetworkGrassrootsReserves2020; @teruiEmergentDualScaling2021]. As suggested by modern conservation initiatives, conservation and management activities should focus on activities that increase the habitat heterogeneity and connectivity of suitable habitat patches to increase resilience to predicted increases in disturbance frequency and magnitude [@schindlerPopulationDiversityPortfolio2010; @schindlerPortfolioConceptEcology2015a; @andersonPortfolioConservationMetapopulations2015].

>>>>

---
title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  pdf_document:
    latex_engine: xelatex
  #word_document:
  #  reference_docx: word_format.docx
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: ecosystems.csl
mainfont: Times New Roman
header-includes:
  \usepackage{textgreek}
---

@@

We employed a discrete-time model to describe tri-trophic dynamics in each habitat patch. A discrete model is suitable for organisms with predictable reproduction cycles, a common life history for many lotic species, including macroinvertebrates and fish. Our simulation model consists of three species; a basal species $B$ with positive logistic growth up to a carrying capacity, $K$; a primary consumer species or intraguild prey, $C$, which grows based on the consumption of $B$; and an intraguild predator $P$, which can consume both $B$ and $C$ (**Figure 1C**). In our model, the realized species abundance $A_{i,x}(t+1)$ for species $i$ (either $B,C,~or~P$) at patch $x$ and time $t+1$ is expressed as a composite of within-patch trophic interactions, immigration, and emigration:

@@

where $s_{0,i}$ is the background survival probability (density independent), $\beta_{i,x}(t)$ the survival probability after predation, and $\gamma_{x}(t)$ the survival probability after stochastic disturbance. The function $g_i(\centerdot)$ defines the form of the density-dependent recruitment for species $i$. Below, we explain how we described predation and recruitment processes for each species.

For the basal species, we considered the predation impacts of consumer and predator species:

@@

where $r_b$ is the reproductive rate for survived individuals.

@@

The interactive effect of disturbance and ecosystem structure has important implications for ecosystem management because anthropogenic climate change is predicted to increase the frequency and magnitude of disturbances [@brunnerFutureStreamflowRegime2020; @mittalImpactHumanIntervention2016]. River networks are especially vulnerable due to their propensity to go dry during droughts [@zipperPervasiveChangesStream2021] as well as susceptibility to more intense flood disturbance with increasing storm intensity [@schiermeierIncreasedFloodRisk2011]. Therefore, the preservation of habitat diversity conferred by ecosystem size and complexity is critical to biodiversity conservation under rapid environmental changes. In addition, maintaining habitat connectivity is particularly important for large, linearly-arranged networks. Although dispersal has received less attention in food chain research [but see @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012a; @wangMetapopulationCapacityDetermines2021], our simulation highlighted food webs in these networks are more vulnerable to habitat fragmentation because habitat patches are inherently isolated from one another. Dams and road crossings have altered the connectivity between suitable habitat patches or have completely blocked dispersal [@ishiyamaPredictingEcologicalImpacts2018; @moritaEffectsHabitatFragmentation2002; @perkinFragmentationAltersStream2012]. Light pollution has also negatively affected the ability of adult aquatic insects to disperse between habitat patches [@perkinEffectsArtificialLighting2014]. These human activities may disproportionately impact food webs in large watersheds with simple network structures.

@@

Our findings add an important distinction to work to assess and prioritize protection or conservation status within human impacted regions. Classical conservation schemes emphasize ecosystem size, aiming to protect the largest areas possible [@diamondIslandDilemmaLessons1975]. While this is certainly a useful attribute generally, designating networks with a high degree of variability, or many small patches, can be more beneficial than a large, homogeneous area [@fahrigWhySeveralSmall2020]. Accounting for ecosystem complexity is particularly suited to ecosystem architectures in which branches have variable environmental conditions [@koningNetworkGrassrootsReserves2020; @teruiEmergentDualScaling2021]. As suggested by modern conservation initiatives, conservation and management activities should focus on activities that increase the habitat heterogeneity and connectivity of suitable habitat patches to increase resilience to predicted increases in disturbance frequency and magnitude [@schindlerPopulationDiversityPortfolio2010; @schindlerPortfolioConceptEcology2015; @andersonPortfolioConservationMetapopulations2015].
